---
output: html_document
editor_options: 
  chunk_output_type: inline
---
Exploration on _Vinho Verde_ Wine by Fabio Tosetto Reale
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Loads all of the required packages

library(ggplot2)
library(gridExtra)

```

```{r echo=FALSE, Load_the_Data}
# Load data from 2 source files into 2 data frames
rw <- read.csv("wineQualityReds.csv")
ww <- read.csv("wineQualityWhites.csv")

# Then, a variant variable is created in each data frame to
  # finally bind them into a single data frame called wine.
rw$variant <- "red"
ww$variant <- "white"
wine <- rbind(rw, ww)

# total.acidity variable is created.
#This is both convenient and makes sense from the perspective of the analysis
wine$variant <- factor(wine$variant)
wine <- transform(wine, total.acidity = fixed.acidity + volatile.acidity)

# Defines conveniently named string variables.
# To be used in plots to color points with intuitive wine colors.
redw.color <- "#a8064a"
whitew.color <- "#eddb8e"
```

The data set explored here contains information on 6497 red and white variants
of _Vinho Verde_ wine. _Vinho Verde_ is a denomination of controlled origin,
which means that all wines in this data set come from the same region in
northern Portugal.
Nonetheless, the denomination is broad enough to contain high variety in
production methods and styles.

The data was separated in two files, one contains data for red wines, the other
for white wines.

These data sets can, of course, be analyzed separetly.
However, I'm interested to see if there are significant diferences between the
red and white variants, both in physicochemical measures and in evaluated
scores.
It's important to observe the data is overrepresented by white wines.

```{r echo=FALSE, summary_1}
table(wine$variant)
```

```{r echo=FALSE, summary_2}
str(wine)
```

```{r echo=FALSE, summary_last_data_transformation}
# Creates a factor variable from the descrete integer variable quality.
# This is convenient for boxplots and color aesthetics.
wine$quality.as.factor <- factor(wine$quality, ordered = TRUE,
                                  levels = seq(3,9,1))

# Conveniently named subsets. One for each variant
rw <- subset(wine, variant == "red")
ww <- subset(wine, variant == "white")

summary(wine)
```

# Univariate Plots Section

```{r echo=FALSE, Univariate_Variant_distribution}
# Simple bar plot for variant
ggplot(wine, aes(variant)) +
  geom_bar()
```

This will be a good visual reminder of how overrepresented white wines are in
this data set.

```{r echo=FALSE, Univariate_Alcohol}
# Histograms of alcohol levels
  # bins = 30 is default behavior. Explicit statement prevents warning
ggplot(wine, aes(alcohol)) +
  geom_histogram(bins = 30) +
  scale_x_continuous(breaks = seq(8,14,1))

# Histogram of alcohol levels faceted by variant
ggplot(wine, aes(alcohol)) +
  geom_histogram(bins = 30) +
  facet_wrap( ~ variant) +
  scale_x_continuous(breaks = seq(8,14,1))
```

I've plotted histograms separate by variant because I've heard many times that
red wines have higher alcohol levels. These plots seem to contradict that, if
the distributions are different it seems whites would be more alcoholic.
But it might be an artifice of the inbalance of the data.
This requires further investigation.

```{r echo=FALSE, Univariate_Sulphates}
# Histograms of sulphates concentration
  # bins = 30 is default behavior. Explicit statement prevents warning
ggplot(wine, aes(sulphates)) +
  geom_histogram(bins = 30)

# Histogram of sulphates concentration faceted by variant
ggplot(wine, aes(sulphates)) +
  geom_histogram(bins = 30) +
  facet_wrap( ~ variant)
```

Sulphates are often said to be the culprit in wine related head aches.
And red wine is said to cause more often and more severe head aches.
There doesn't seem to be significant differences in the distributions, but I'll
investigate this a little more.

```{r echo=FALSE, Univariate_Density}
# Histograms of density
  ggplot(wine, aes(density)) +
  geom_histogram(binwidth = 0.0005)

# Histogram for density faceted by variant
ggplot(wine, aes(density)) +
  geom_histogram(binwidth = 0.0005) +
  facet_wrap( ~ variant)
```

Density seems to be bimodal.
The faceted plots indicate there is likely a difference in distributions for
red and white wines.

```{r echo=FALSE, Univariate_Acidity_profile}
# Histograms for acidity related variables:
# pH, total.acidity, fixed.acidity and citric acid
  # bins = 30 is default behavior. Explicit statement prevents warning
ggplot(wine, aes(pH)) +
  geom_histogram(bins = 50)

ggplot(data = wine, aes(total.acidity)) +
  geom_histogram(bins = 30)

ggplot(data = wine, aes(fixed.acidity)) +
  geom_histogram(bins = 30)

ggplot(data = wine, aes(volatile.acidity)) +
  geom_histogram(bins = 30)

ggplot(data = wine, aes(citric.acid)) +
  geom_histogram(bins = 30)
```

All of the acidity variables seem to somewhat follow a normal distribution.
There is one possible exception with _volatile.acidity_.
There might still be some interesting relationship between these variables.


```{r echo=FALSE, Univariate_Sugar}
# Histograms for residual.sugar
  # bins = 30 is default behavior. Explicit statement prevents warning
ggplot(wine, aes(residual.sugar)) +
  geom_histogram(bins = 30)

# Histogram for residual.sugar faceted by variant
ggplot(wine, aes(residual.sugar)) +
  geom_histogram(bins = 30) +
  facet_wrap( ~ variant)
```

Interesting, I expected _residual.sugar_ to inversely correlate strongly with 
_alcohol_ since the sugar is converted to alcohol in the fermentation process.
But that doesn't seem likely anymore, due to how different the histograms look.

Also, dry wines are likely the ones with the least amounts of _residual.sugar_, 
so it looks like white wines are more likely sweet then red wines.

```{r echo=FALSE, Univariate_Chlorides}
# Histogram for chlorides
  # bins = 30 is default behavior. Explicit statement prevents warning
ggplot(wine, aes(chlorides)) +
  geom_histogram(bins = 30)
```

```{r echo=FALSE, Univariate_Sulfur}
# Histograms for sulfur dioxide variables:
# free.sulfur.dioxide, total.sulfur.dioxide
  # bins = 30 is default behavior. Explicit statement prevents warning
ggplot(wine, aes(free.sulfur.dioxide)) +
  geom_histogram(bins = 30)

ggplot(wine, aes(total.sulfur.dioxide)) +
  geom_histogram(bins = 30)
```

It seems _total.sulfur.dioxide_ is bimodal.
Like _density_, this might be a distinguishing feature for red vs white wines.

```{r echo=FALSE, Univariate_Sulfur2}
# Histogram for total.sulfur.dioxide faceted by variant
  # bins = 30 is default behavior. Explicit statement prevents warning
ggplot(wine, aes(total.sulfur.dioxide)) +
  geom_histogram(bins = 30) +
  facet_wrap( ~ variant)
```

```{r echo=FALSE, Univariate_Quality}
# Bar plots for quality. Factor version of the variable used for convenience
ggplot(wine, aes(quality.as.factor)) +
  geom_bar() +
  xlab("quality")

# Bar plot for quality faceted by variant
ggplot(wine, aes(quality.as.factor)) +
  geom_bar() +
  facet_wrap( ~ variant) +
  xlab("quality")
```

_Quality_ is overwhelmingly 5 or 6 for both wines. The side by side plot seem
to indicate that white wines are slightly better evaluated.

# Univariate Analysis


### What is the structure of your dataset?

The data set consists of 6497 observations of physicochemical measures and
subjective sensory avaluations.
There is also a factor variable describing if any given observation is from a
red or white wine.

The variables are:
 
 * 11 quantitative mesures, like _alcohol_ (in % by volume) or _pH_.
 
 * 1 numerical qualitative sensory evaluation. Given as the median of at least
 3 evaluations made by wine experts. Scale ranges from 0 (_bad_) to 10 (_good_).
 
 * 1 indexing variable.
 
There are also 2 variables added:
 
 * 1 categorical variable identifying _red_ or _white_ wine.
 
 * 1 agregate variable _total.acidity_ and was created beacause this information
 is usually provided on labels.

Most variables seem to follow a bell-shaped distribution, but not necessarily
Gaussian.
The _alcohol_ variable seem to have a more spread out distribution while
_residual.sugar_ appears to be left skewed.
Also, _total.sulfur.dioxide_ looks bimodal on agregate, but it seems to be 
explained by a difference in distributions for red and white wines.

Surprisingly, so far, red and white wines seem to follow the same distribution
on all of the investigated variables, with the exeption of _total.sulfur.oxide_
and, possibly, _alcohol_ and _quality_.
All of these exceptions appear to be skewed in favor of larger values for the 
white variaty. This goes against my intuition in the cases of _alcohol_ and
_quality_

### What is/are the main feature(s) of interest in your dataset?

I'm interested in the _quality_ score, more precisely, I'm interested to see
how this variable relates to physicochemical ones, specially those contained
in labels.

I'm also interested in the _variant_ variable. Specifically, I'm curious if the
physicochemical variables could indicate subtle categorical differences in the
variants.
Regarding the variants, I'm also interested to see if there is a bias in the
expert community in evaluating these wines.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Wine experts like to say the 4 most import features of wine are _alcohol_,
_tannin_, _sweetness_ and _acid_ profile.
The data set has a direct _alcohol_ variable, so that looks promising.
The _sweetness_ feature is a function of the _residual.sugar_ variable.
Acid profile is trickier.
There is the obvious _pH_ variable, but flavors come with which type of acid is
present and in which concentration.
The variables _fixed.acidity_, _volatile.acidity_, and _citric.acid_ should
help.
Unfortunately, none of the variables are related to _tannin_ as far as I could 
find. These are supposed to be complex organic molecules and don't seem to 
relate to any of the chemical quantities described in the data.

### Did you create any new variables from existing variables in the dataset?

Yes, I created _total.acidity_. It is just the sum of 2 other variables,
_fixed.acidity_ and _volatile.acidity_ but the interest in this particular
quantity is due to the fact that this is the information most universaly
contained in wine labels.

# Bivariate Plots Section

```{r echo=FALSE, Bivariate_Quality_Alcohol}
# Scatter plots for alcohol and quality
# Points are jittered to deal with overplotting,
  #but only for quality (since it is discrete)
ggplot(data = wine, aes(quality, alcohol)) +
  geom_jitter(alpha = 0.25, height = 0) +
  scale_x_continuous(breaks = seq(3,9,1)) +
  scale_y_continuous(breaks = seq(8,15,1))

# Scatter plot for alcohol and quality faceted by variant
ggplot(data = wine, aes(quality, alcohol)) +
  geom_jitter(alpha = 0.25, height = 0) +
  scale_x_continuous(breaks = seq(3,9,1)) +
  scale_y_continuous(breaks = seq(8,15,1)) +
  facet_wrap( ~ variant)
```

Looks like the higer _quality_ wines are more alcoholic.
Let's see if a box plot looks better.

```{r echo=FALSE, Bivariate_Quality_Alcohol_2}
# Box plots for alcohol and quality
ggplot(wine, aes(quality.as.factor, alcohol)) +
  geom_boxplot() +
  xlab("quality")

# BOx plot for alcohol and quality faceted by variant
ggplot(wine, aes(quality.as.factor, alcohol)) +
  geom_boxplot() +
  xlab("quality") +
  facet_wrap( ~ variant)
```

The median level of _alcohol_ is clearly increasing with score.
There are quite a few outliers with high _alcohol_ level and _quality_ score of
5, though.

Also, the distribution for _alcohol_ now seems to be the same for red and white
wines, so I'll just plot a qq-plot.

```{r echo=FALSE, QQ_Alcohol}
# Plots a qq plot where the points are the percentiles
  # percent is a vector with the probability to pass in the quantile function.
# Then, geom_point must get these points in the aes() wrapper.
# Notice that rw and ww are named subsets of wine containing, respectively,
  # only red or white wines.
# Finally, we call geom_abline to plot a dashed y = x line.
percent <- seq(0.01, 1.0, 0.01)
ggplot() +
  geom_point(aes(x = quantile(rw$alcohol, percent),
                 y = quantile(ww$alcohol, percent))) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  xlab("alcohol (reds)") +
  ylab("alcohol (whites)")
```

The distributions are close enough to one another to be considered the same.

```{r echo=FALSE, Bivariate_Quality_Acidity}
# Box plots for alcohol and quality
ggplot(wine, aes(quality.as.factor, total.acidity)) +
  geom_boxplot() +
  xlab("quality")

# Box plot for alcohol and quality faceted by variant
ggplot(wine, aes(quality.as.factor, total.acidity)) +
  geom_boxplot() +
  xlab("quality") +
  facet_wrap( ~ variant)
```

Interquartile range seems to be *decreasing* whith the increase in _quality_ in
the agregate, but *increasing* for each _variant_.

There also seem to be some difference in the acidity distributions of red and
white wines.
I'll plot a qq-plot to find out.

```{r echo=FALSE, QQ_Acidity}
# Plots a qq plot where the points are the percentiles
  # percent is a vector with the probability to pass in the quantile function,
  # and was defined previously in the code.
# Then, geom_point must get these points in the aes() wrapper.
# Notice that rw and ww are named subsets of wine containing, respectively,
  # only red or white wines.
# Finally, we call geom_abline to plot a dashed y = x line.
ggplot() +
  geom_point(aes(x = quantile(rw$total.acidity, percent),
                 y = quantile(ww$total.acidity, percent))) +
  geom_abline(intercept=0, slope=1, linetype = "dashed") +
  xlab("total acidity (reds)") +
  ylab("total acidity (whites)")
```

Alright, that settles it. Red wines are definetly more acidic.

Now, getting back to quality, let's see if _citric.acid_ gives us some insight.

```{r echo=FALSE}
# Box plots for citric.acid and quality
ggplot(wine, aes(quality.as.factor, citric.acid)) +
  geom_boxplot() +
  xlab("quality")

# Box plot for citric.acid and quality faceted by variant
ggplot(wine, aes(quality.as.factor, citric.acid)) +
  geom_boxplot() +
  xlab("quality") +
  facet_wrap( ~ variant)
```

Now, that is interesting! The _citric.acid_ variable seems to predict _quality_
for red wines only.

```{r echo=FALSE, warning=FALSE, Bivariate_Quality_Sugar}
# Box plots for residual.sugar and quality
# Data passed is a subset of wine.
  # This was done to remove an outlier from the plot
ggplot(subset(wine, residual.sugar < 40),
       aes(quality.as.factor, residual.sugar)) +
  geom_boxplot() +
  xlab("quality")

# Box plot for residual.sugar and quality faceted by variant
ggplot(subset(wine, residual.sugar < 40),
       aes(quality.as.factor, residual.sugar)) +
  geom_boxplot() +
  facet_wrap( ~ variant) +
  xlab("quality")
```

In all of the above boxplots there is an omission, a white wine outlier with 
65.8 g/L in _residual sugar_.

```{r echo=FALSE}
# Creates a subset containing the points omited in the above plot
  # only a few variables are selected for ease of visualization
subset(wine, residual.sugar > 40,
       select = c("quality", "variant","residual.sugar", "alcohol",
                  "density"))
```

We can observe the median values go up and down with _quality_ due to the white
wines.
Red wines have a veri consistent median across all _quality_ values, which was
to be expected since most _residual.sugar_ values are in a very narrow band of
values.
Also, white wines are largely sweeter then reds in their distributions, no
qq-plot required for this observation.

```{r echo=FALSE, warning=FALSE, Bivariate_Quality_Density}
# Box plots for residual.sugar and quality
# Data passed is a subset of wine.
  # This was done to remove an outlier from the plot. The data is subset by
    # residual.sugar because it is the same outlier identified before
ggplot(subset(wine, residual.sugar < 40),
       aes(quality.as.factor, density)) +
  geom_boxplot() +
  xlab("quality")

# Box plot for residual.sugar and quality faceted by variant
ggplot(subset(wine, residual.sugar < 40),
       aes(quality.as.factor, density)) +
  geom_boxplot() +
  facet_wrap( ~ variant) +
  xlab("quality")
```

Quality seems to be inversely correlated with quality. This relationship appear
stronger for white wines.
If memory serves me right, alcohol has a lower density then water. So, this 
relationship might be superfluous.
Let's see how related are they.

```{r echo=FALSE, warning=FALSE, Bivariate_Alcohol_Density}
# Scatter plot for density and alcohol
# Data passed is a subset of wine.
  # This was done to remove an outlier from the plot. The data is subset by
    # residual.sugar because it is the same outlier identified before
ggplot(subset(wine, residual.sugar < 40),
       aes(alcohol, density)) +
  geom_point(alpha=0.05)
```

In all of the density plots there is an omission, the same white wine outlier 
with 65.8 g/L in _residual sugar_, which has a _density_ of 1.03898 g/L.

Let's check if _residual.sugar_ explains density well.

```{r echo=FALSE, warning=FALSE, Bivariate_Sugar_Density}
# Scatter plot for density and residual.sugar
# Data passed is a subset of wine.
  # This was done to remove an outlier from the plot. The data is subset by
    # residual.sugar because it is the same outlier identified before
ggplot(subset(wine, residual.sugar < 40),
       aes(residual.sugar, density)) +
  geom_point(alpha=0.05)
```

These are certainly 2 distributions plotted together.
Most likely, this is a distinguishing feature for red and white wines.
Let's see, subseting on wines with less than 20 g/L in _residual.sugar_

```{r echo=FALSE, warning=FALSE, Bivariate_Sugar_Density2}
# Scatter plot for density and residual.sugar faceted by variant
# Data is subset to zoom in on lower values
ggplot(subset(wine, residual.sugar < 20),
       aes(residual.sugar, density)) +
  geom_point(alpha=0.05) +
  scale_x_continuous(breaks = seq(0,30,2.5)) +
  facet_wrap( ~ variant)
```

Interesting, white wines seem to have 2 distinct distributions in respect to
density.
One of these behaviors seem to follow closely what happens with reds, but 
in lower densities.


```{r echo=FALSE, Bivariate_Quality_Sulphates}
# Box plots for sulphates and quality
ggplot(wine,aes(quality.as.factor,sulphates)) +
  geom_boxplot() +
  xlab("quality")

# Box plots for sulphates and quality faceted by variant
ggplot(wine,aes(quality.as.factor,sulphates)) +
  geom_boxplot() +
  facet_wrap(~variant) +
  xlab("quality")
```

Much like _citric.acid_, median _sulphates_ values increase with _quality_.
Let's see if there is some interesting relationship between them.

```{r echo=FALSE, Bivariate_Red_Citric_Sulphates}
# Scatter plots for citric.acid and sulphates
ggplot(wine, aes(citric.acid, sulphates)) +
  geom_point(alpha=0.1)

# Scatter plot for citric.acid and sulphates faceted by variant
ggplot(wine, aes(citric.acid, sulphates)) +
  geom_point(alpha=0.1) +
  facet_wrap( ~ variant)
```

Disappointing! I'll investigate in more detail later.

```{r echo=FALSE, Bivariate_Acidity_Alcohol}
# Scatter plots for alcohol and total.acidity
ggplot(wine, aes(alcohol, total.acidity)) +
  geom_point(alpha = 0.1)

# Scatter plots for alcohol and acidity variables faceted by variant:
  # total.acidity and pH
ggplot(wine, aes(alcohol, total.acidity)) +
  geom_point(alpha = 0.1) +
  facet_wrap( ~ variant)

ggplot(wine, aes(alcohol, pH)) +
  geom_point(alpha = 0.1) +
  facet_wrap( ~ variant)
```

Experts advocate for a balance in alcohol and acidity.
I would expect from that for _alcohol_ and _total.acidity_ to show some sort of
correlate tendency.
The fact they don't might explain why most wines are not well evaluated.

```{r echo=FALSE, Bivariate_Acidities}
# Scatter plots for fixed.acidity and volatile.acidity
ggplot(wine, aes(fixed.acidity, volatile.acidity)) +
  geom_point(alpha = 0.1)

# Scatter plot for fixed.acidity and volatile.acidity faceted by variant
ggplot(wine, aes(fixed.acidity, volatile.acidity)) +
  geom_point(alpha = 0.1) +
  facet_wrap( ~ variant)
```

As bivariate distribution goes, red wines have more variance within both acidity
measures.

```{r echo=FALSE, Bivariate_Fixed_Citrics}
# Scatter plots for citric.acid and fixed.acidity
ggplot(wine, aes(fixed.acidity, citric.acid)) +
  geom_point(alpha = 0.1)

# Scatter plots for citric.acid and fixed.acidity faceted by variant
ggplot(wine, aes(fixed.acidity, citric.acid)) +
  geom_point(alpha = 0.1) +
  facet_wrap( ~ variant)
```

White wines are contained in narrow bands in the acidity categories.
For red wines, _citric.acid_ and _fixed.acid_ are well correlated.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Overall, only _alcohol_ and _density_ seems to have any kind of significant
relationship with _quality_.
However, these 2 variables have significant correlation with each others.
These overall observations seem to be the only ones aplicable for white wines,
which are overrepresented in the data set.
It is worth mentioning that the relationship between _density_ and _quality_
is stronger for white wines.
Every other interesting relationship with quality apply well enough only for
red wines, and these are between _quality_ and _alcohol_, _sulphates_ and 
_citric.acid_.

As for distinguishing the _variant_ variable, in adition to _sulphates_ and 
_citric.acid_, _residual.sugar_ are likely distributed differently between
variants.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Yes. The variables _density_ and _residual.sugar_ seem to have distinct 
distributions for red and white wines. More over, there seems to be 2 distinct 
relationships between these variables within white wines.
This indicates there is a hidden categorical variable in the white wines 
category.

### What was the strongest relationship you found?

Overall, the relationship of _alcohol_ and _quality_ was the strongest.
When restricting to red wines, I would say _citric.acid_ and _quality_ showed
the strongest relationship.

# Multivariate Plots Section


```{r echo=FALSE, warning=FALSE, Multivariate_Sugar_Density}
# Scatter plot of residual.sugar and density
  # faceted by variant and colored by quality
# Once again, data is subset to exclude one outlier
ggplot(subset(wine, residual.sugar < 40),
       aes(residual.sugar, density)) +
  geom_point(alpha = 0.5, aes(color = quality.as.factor)) +
  facet_wrap( ~ variant)
```

Starting with the curious relationship between _density_ and _residual.sugar_.
It seems that, for white wines and fixed density, an increase in residual.sugar
is related with an increase in _quality_.

Let's say wines with less then 3 g/l in _residual.sugar_ are *dry* wines and the
rest are *sweet* wines. Let's check that we can distinguish red from white *dry*
wines.

```{r echo=FALSE, Multivariate_Sugar_Density2}
# Scatter plot of residual.sugar and density colored by variant
# Data is subset to zoom in the low end of residual.sugar levels
# Points are jittered to deal with overplotting
# Points are colored in intuitive wine colors and background is made darker
  # to interact bertter with the chosen wine colors
ggplot(subset(wine, residual.sugar < 3),
       aes(residual.sugar, density)) +
  geom_jitter(height = 0, aes(color = variant)) +
  scale_color_manual(values=c(redw.color, whitew.color)) +
  theme(panel.background = element_rect(fill = "grey"))

# Scatter plot of residual sugar and density colored by quality
ggplot(subset(wine, residual.sugar < 3),
       aes(residual.sugar, density)) +
  geom_jitter(height = 0, aes(color = quality.as.factor)) +
  theme(panel.background = element_rect(fill = "grey"))
```

Interesting, the relationship with _quality_ observed for white wines seems to
exist for the *dry* wines as well.

```{r echo=FALSE, Multivariate_Sugar_Density3}
# Scatter plot of residual sugar and density colored by quality
# Data is subset to zoom on higher residual.sugar values, but avoiding outliers
ggplot(subset(wine, residual.sugar > 3 & residual.sugar < 25),
       aes(residual.sugar, density)) +
  geom_point(alpha = 0.8, aes(color = quality.as.factor)) +
  theme(panel.background = element_rect(fill = "grey"))
```

When restricting to *sweet* wines, this low _density_ higher _residual.sugar_ 
relationship looks stronger.

```{r echo=FALSE, Density_Sugar_Combinations}
# Eye-balled angular coeficients for somewhat constant linear relationships
  # for density and residual.sugar
# There are 2 distinct relationships made, one for dry and one for sweet wines
# 1000*density can be understood as converting density to g/L which is the unit
  # used elsewhere in the data.
wine <- transform(wine,
                  dry.density.sugar = 1000*density - 4*residual.sugar)
wine <- transform(wine,
                  sweet.density.sugar = 1000*density - 0.533*residual.sugar)

# Scatter plot of alcohol and the linear combination for dry wines
  # faceted by variant and colored by quality
# Data is subset to zoom in on dry wines
ggplot(subset(wine, residual.sugar < 3),
              aes(alcohol, dry.density.sugar)) +
  geom_point(aes(color = quality.as.factor)) +
  facet_wrap( ~ variant) +
  theme(panel.background = element_rect(fill = "grey"))

# Scatter plot of alcohol and the linear combination for dry wines
  # colored by quality
# Data is subset to zoom in on sweet wines but avoiding outliers
ggplot(subset(wine, residual.sugar>3 & residual.sugar<25 & variant=="white"),
              aes(alcohol, sweet.density.sugar)) +
  geom_point(aes(color = quality.as.factor)) +
  theme(panel.background = element_rect(fill = "grey"))
```

Notice that, in the above plots, the y-axis measures are different linear
combinations of the _density_ and _residual.sugar_.

```{r echo=FALSE}
# Scatter plot for citric.acid and alcohol colored by quality
ggplot(wine, aes(alcohol, citric.acid)) +
  geom_point(aes(color = quality.as.factor)) +
  theme(panel.background = element_rect(fill = "grey"))
```

Signals found to discriminate _quality_ seem to work well (linearly), together.
To build a predictive model for _quality_, I would recomend a hybrid aproach.
The first step would be to use the bimodal distributions like _residual.sugar_
to create a *dry-red*, *dry-white* or *sweet* classifier.
Next, a different model for each, which should benefit from dimensionality
reduction (_PCA_ looks good as relations seem linear).

```{r echo=FALSE, Citric_Sulphate_Blobs}
# Scatter plot for citric.acid and sulphates
  # faceted by variant and colored by quality
# Data is subset to avoid high citric.acid outliers
ggplot(subset(wine, citric.acid<1), aes(citric.acid, sulphates)) +
  geom_point(aes(color = quality.as.factor)) +
  facet_wrap( ~ variant) +
  theme(panel.background = element_rect(fill = "grey"))

# Scatter plot for citric.acid and sulphates colored by quality
# Data is subset to remove 5 and 6 quality wines
ggplot(subset(rw, quality < 5 | quality > 6), aes(citric.acid, sulphates)) +
  geom_point(aes(color = quality.as.factor)) +
  theme(panel.background = element_rect(fill = "grey"))
```

This an example of other feature relations. We can observe that the originaly 
observed tendency to distinguish _quality_ is mostly contained in one direction.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

In both of the features of interest, _residual.sugar_ and _density_ worked
together to strengthen the signal of good vs. bad _quality_ wines and
distinguishing red vs. white _variants_.

### Were there any interesting or surprising interactions between features?

Yes. There are 2 distinct behaviors in the relation between _density_ and 
_residual.sugar_. I believe these distinct behaviors distinguish dry and sweet
wines.
The surprising relation is that, for dry wines of some fixed _density_, an
*increase* in _residual.sugar_ is associated with an *increase* in _quality_.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
# Scatter plot for density and residual.sugar
# Data is subset to remove high residual.sugar outliers
# Added a first aproximation line to use in classifying dry vs sweet wines
ggplot(subset(wine, residual.sugar < 25),
       aes(residual.sugar, density)) +
  geom_point(alpha = 0.1) +
  scale_x_continuous(breaks = seq(0, 25, 2.5)) +
  xlab("Residual Sugar (g/L)") +
  ylab("Density (Kg/L)") +
  geom_abline(slope = 1/125, intercept = 121/125,
              color = "red", linetype = "dashed")
```

### Description One

The relationship of _Density_ and _Residual Sugar_ appear to follow 2 distinct 
trends. This is likely a difference in composition of dry vs. sweet wines.

### Plot Two
```{r echo=FALSE, Plot_Two}
# Scatter plot for residual.sugar and density faceted by variant
# Data is subset to remove high residual.sugar outliers
# Added a first aproximation line to use in classifying red vs white wines
ggplot(subset(wine, residual.sugar < 25),
       aes(residual.sugar, density)) +
  geom_point(alpha = 0.5, aes(color = variant)) +
  facet_wrap( ~ variant) +
  scale_color_manual(values=c(redw.color, whitew.color)) +
  scale_x_continuous(breaks = seq(0, 24, 3)) +
  xlab("Residual Sugar (g/L)") +
  ylab("Density (Kg/L)") +
  theme(panel.background = element_rect(fill = "grey")) +
  geom_abline(slope = 1/2000, intercept = 993.25/1000,
              color = "purple", linetype = "dashed")
```

### Description Two

The relationship of _Density_ and _Residual Sugar_ also appear to allow for a 
linear model to make good predictions on classifying red from white wine
_variants_.

### Plot Three
```{r echo=FALSE, Plot_Three}
# These are mostly the same plots in the Density_Sugar_Combination chunk
  # with some adjustments to polish presentation
pl1 <- ggplot(subset(wine, residual.sugar < 3),
                     aes(alcohol, dry.density.sugar)) +
         geom_point(aes(color = quality.as.factor)) +
         xlab("") +
         ylab("Density - 4 * Res. Sugar\n (g/L)") +
         facet_wrap( ~ variant) +
         theme(panel.background = element_rect(fill = "grey"),
               legend.position = "none")

pl2 <- ggplot(subset(wine,
                     residual.sugar>3 & residual.sugar<25 & variant=="white"),
                     aes(alcohol, sweet.density.sugar)) +
         geom_point(aes(color = quality.as.factor)) +
         xlab("Alcohol (% volume)") +
         ylab("Density - 0.533 * Res. Sugar\n (g/L)") +
         theme(panel.background = element_rect(fill = "grey"),
               legend.title=element_blank())

grid.arrange(pl1, pl2, ncol = 1)
```

### Description Three

For dry vs. sweet wines, different linear combinations of _Density_ and 
_Residual Sugar_ seem to be usefull in predicting good wines.
Moreover, for dry wines, red and white _variants_ seem to have somewhat distinct
characteristics for predicting good wines.

------

# Reflection

The _Vinho Verde_ data set consists of 6497 observations of physicochemical 
measures and subjective sensory avaluations. The set is unbalanced, containing
3 times more white wines than red wines (4898 vs. 1599).

There are only 13 variables, which made possible to avaluate most of them 
indivualy, before looking for relationships within variables.
In this first exploration, some variables had interesting behaviors.
_Alcohol_ had the more spead out variance.
Some variables looked bimodal, which made for interesting explorations in the
potential differences of red and white wines.
_Density_ ended up being the most interesting of these variables.

In exploring _quality_ and how that relates to other variables, _alcohol_ was 
the clear signal of good wines in agregate.
We ended up finding the unbalanced data were hiding some _quality_ identifying
variables, as red wines had more variables relating to _quality_.

But, in the end, it became appearant there were 3 distinct categories of  wine, 
for which the physicochemical variables held different relations to _quality_.

With the above realization in mind, future work aiming to create predictors for
good wines should focus on a strategy to, in a first step, trying to classify
wines in one of 3 categories: *red*, *dry-white* or *sweet-white*.
Both in identifying _red_ from _white_ or _dry_ from _sweet_, _density_ and
_residual sugar_ are usefull variables.
There are some variables which seem promising in the identification of _red_
wines from _white_ ones.
These include _total.sulfur.dioxide_, _total.acidity_ and _sulphates_.
A different, specific model can be made for each of these 3 classified wines.
_Alcohol_ is relevant for all of these models.
For _white_ wines (both dry and sweet), _density_ and _residual sugar_ seem to be the best variables to include in the model (after _alcohol_).
For _red_ wines, on the other hand, other variables seem best, like _sulphates_
or _citric acid_.

The most relevant limitation of the data set is the _Denomination of Origin_.
All of the wines in the data set are from the same region in Portugal.
The results are likely not generalizable.
It was somewhat troublesome to deal with the unbalance in the variaties.
This was specially the case since red wines had more variables signalling
good wines.
When taking the agregate data these relationships were masked.

------